---
title: "Introduction to eegwhiten"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to eegwhiten}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`eegwhiten` provides whitening transforms for EEG and multichannel signal data.
It supports six methods (PCA, PCA-cor, ZCA, ZCA-cor, SVD, Cholesky) with a
model-based API: fit once on training data, then apply to new data.

## Quick Start

```{r quick-start}
library(eegwhiten)
set.seed(42)

# Simulate EEG data: 300 trials x 16 channels
X_train <- matrix(rnorm(300 * 16), 300, 16)
X_test  <- matrix(rnorm(120 * 16), 120, 16)
colnames(X_train) <- paste0("Ch", 1:16)
colnames(X_test)  <- colnames(X_train)

# Fit a whitening model on training data
m <- whiten_model(
  X_train,
  method       = "PCA",
  var_threshold = 0.95,
  lambda       = "auto"
)
print(m)

# Apply to test data
Z_test <- predict(m, X_test)
dim(Z_test)
```

## Diagnostics

After whitening, the covariance of the transformed data should be close to
the identity matrix. Use `check_whitening()` to verify:

```{r diagnostics}
d <- check_whitening(Z_test)
d$diag_mean       # should be close to 1
d$offdiag_frob    # should be close to 0
```

## Inverse Transform

Recover an approximation of the original data from whitened data:

```{r inverse}
X_rec <- unwhiten(Z_test, m)
dim(X_rec)
```

For full-rank methods (no dimensionality reduction), the reconstruction is exact.

## One-Shot Whitening

For simple workflows where you don't need the model object:

```{r one-shot}
res <- whiten_matrix(X_train, method = "ZCA", lambda = 0.1)
str(res$Z)
```

## Whitening Methods

| Method    | Description                          | Dim. reduction |
|-----------|--------------------------------------|:--------------:|
| PCA       | Eigen-decomposition based            | Yes            |
| PCA-cor   | PCA on correlation (scale-invariant) | Yes            |
| SVD       | Singular value decomposition         | Yes            |
| ZCA       | Zero-phase, closest to original data | No             |
| ZCA-cor   | ZCA on correlation                   | No             |
| Cholesky  | Cholesky factorization               | No             |

## Regularization

Covariance shrinkage toward a scaled identity target prevents
ill-conditioning:
- `lambda = 0`: no shrinkage
- `lambda = "auto"`: automatic selection via OAS or Ledoit-Wolf
- `lambda` between 0 and 1: manual shrinkage

```{r regularization}
m_reg <- whiten_model(X_train, method = "PCA", lambda = 0.2)
m_auto <- whiten_model(X_train, method = "PCA", lambda = "auto",
                       lambda_method = "oas")
m_auto$lambda
```

## Batch Whitening

Process multiple sessions or subjects at once:

```{r batch}
X_list <- list(
  matrix(rnorm(200 * 8), 200, 8),
  matrix(rnorm(180 * 8), 180, 8)
)

# Shared model from first matrix
out <- whiten_batch(X_list, mode = "shared_model",
                    method = "PCA", n_comp = 4, lambda = 0.1)
length(out)
```

## Euclidean Alignment

Align multiple sessions to a global reference covariance:

```{r ea}
ea <- euclidean_alignment(X_list, input = "raw",
                          mean_method = "logeuclid")
str(ea$aligned, max.level = 1)
```

## Incremental Updates

Update a model with new data without refitting from scratch:

```{r update}
X_new <- matrix(rnorm(80 * 16), 80, 16)
m_updated <- whiten_model_update(m, X_new)
m_updated$n_obs
```

## Hyperparameter Tuning

Automatically select the best method and parameters via cross-validation:

```{r tuning}
tuned <- auto_tune_whitening(
  X_train,
  methods      = c("PCA", "ZCA"),
  n_comp_grid  = c(8, 12),
  lambda_grid  = list("auto", 0.1),
  cv_folds     = 3,
  seed         = 123,
  top_n        = 4
)
head(tuned$ranking)
```

## Tensor Data

Handle 3D EEG tensors (trials x channels x time):

```{r tensor}
X_tensor <- array(rnorm(20 * 8 * 50), dim = c(20, 8, 50))
m_t <- whiten_model_tensor(X_tensor, method = "ZCA", lambda = 0)
Z_tensor <- predict_tensor(m_t, X_tensor)
dim(Z_tensor)
```

## Reporting

Generate a markdown report:

```{r report}
txt <- report_whitening(m, data = X_train, file = NULL)
cat(substr(txt, 1, 300), "...\n")
```
