---
title: "Introduction to eegwhiten"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to eegwhiten}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.4,
  dpi = 110
)
set.seed(42)

# small helper: heatmap of a matrix with a diverging palette
heat <- function(M, main, zlim = c(-1, 1)) {
  d <- ncol(M)
  pal <- grDevices::colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(64)
  graphics::image(seq_len(d), seq_len(d), t(M[d:1, , drop = FALSE]),
                  zlim = zlim, col = pal, axes = FALSE, xlab = "", ylab = "", main = main)
  graphics::box(col = "grey70")
}
```

## Overview

`eegwhiten` provides covariance whitening and covariance-based transfer-learning
tools for EEG and other multichannel signals. The core is a model-based
fit/transform workflow:

- **Rows** are trials/epochs/samples; **columns** are channels/features.
- Fit once on training data, then apply to new data.
- Six whitening methods, covariance shrinkage, robust estimators, dimensionality
  reduction, and cross-validated tuning.

On top of whitening it adds the pieces needed for cross-session / cross-subject
work: **Euclidean alignment** and **per-domain recentering**, **Riemannian
tangent-space** mapping, and **generalized-eigenvalue ("relative") whitening**.

```{r library}
library(eegwhiten)
```

## Quick start: fit, transform, invert

```{r quick-start}
X_train <- matrix(rnorm(300 * 16), 300, 16)
X_test  <- matrix(rnorm(120 * 16), 120, 16)
colnames(X_train) <- paste0("Ch", 1:16)
colnames(X_test)  <- colnames(X_train)

m <- whiten_model(X_train, method = "PCA", var_threshold = 0.95, lambda = "auto")
print(m)

Z_test <- predict(m, X_test)
dim(Z_test)

# Inverse transform (note: model-first argument order)
X_rec <- unwhiten(m, Z_test)
dim(X_rec)
```

## Diagnostics

After whitening, `cov(Z)` should be close to the identity. `check_whitening()`
reports several measures, including a bounded `whiteness` score in `[0, 1]`.

```{r diagnostics}
d <- check_whitening(predict(m, X_train))
c(diag_mean = d$diag_mean, offdiag_frob = d$offdiag_frob,
  whiteness = d$whiteness, logdet = d$logdet)
```

Visually, the strong off-diagonal correlation structure of the input is removed:

```{r vis-before-after, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1))
heat(cor(X_test), "before: cor(X)")
heat(cov(predict(m, X_test)), "after: cov(Z)")
```

## Whitening methods

| Method    | Description                          | Dim. reduction |
|-----------|--------------------------------------|:--------------:|
| PCA       | Eigen-decomposition based            | Yes            |
| PCA-cor   | PCA on correlation (scale-invariant) | Yes            |
| SVD       | Singular value decomposition         | Yes            |
| ZCA       | Zero-phase, closest to original data | No             |
| ZCA-cor   | ZCA on correlation                   | No             |
| Cholesky  | Cholesky factorization               | No             |

## Regularization and shrinkage targets

Covariance shrinkage stabilizes whitening for high-dimensional / low-sample
data. `lambda = "auto"` chooses the intensity with OAS or Ledoit-Wolf. The
target is a scaled identity by default, or `"diagonal"` to shrink correlations
toward zero while preserving each channel's variance.

```{r shrinkage}
m_auto <- whiten_model(X_train, method = "ZCA", lambda = "auto", lambda_method = "oas")
m_auto$lambda

m_diag <- whiten_model(X_train, method = "ZCA", lambda = 0.3, shrink_target = "diagonal")
m_diag$shrink_target
```

Robust covariance estimators (`"tyler"`, and `"mcd"` with **robustbase**) are
available via `cov_estimator`.

## Spatial filters and patterns

The whitening filters (rows of `W`) and the corresponding forward patterns are
both available. Patterns -- not raw filter weights -- are the right object to
interpret as scalp topographies (Haufe et al., 2014).

```{r patterns}
fp <- whitening_patterns(whiten_model(X_train, method = "PCA", n_comp = 4))
dim(fp$filters)   # components x channels
dim(fp$patterns)  # channels x components
```

## Hyperparameter tuning

`auto_tune_whitening()` cross-validates over methods and parameters and returns
a tuned object you can predict with directly.

```{r tuning}
tuned <- auto_tune_whitening(
  X_train, methods = c("PCA", "ZCA"),
  n_comp_grid = c(8, 12), lambda_grid = list("auto", 0.1),
  cv_folds = 3, seed = 123, top_n = 4
)
print(tuned)

Z <- predict(tuned, X_test)   # uses the best model
dim(Z)
```

## Cross-session / cross-subject transfer

Independent per-matrix whitening removes the very between-session covariance
structure transfer learning needs. `eegwhiten` offers alignment instead.

**Per-domain recentering** (`recenter()`) aligns each session by its *own*
covariance, mapping every domain to the identity -- the Euclidean Alignment idea
of He and Wu (2019):

```{r recenter}
X_list <- list(
  matrix(rnorm(200 * 8), 200, 8),
  matrix(rnorm(180 * 8), 180, 8) %*% diag(1:8)  # a shifted session
)

out <- whiten_batch(X_list, mode = "recenter")
round(diag(cov(out[[2]]$Z)), 3)   # each domain -> identity
```

Across several drifting sessions, recentering collapses the between-session
Riemannian distances toward zero:

```{r vis-align, fig.height = 3.4, fig.width = 6}
sessions <- lapply(1:6, function(s) {
  S <- crossprod(matrix(rnorm(64), 8, 8)) + diag(8)
  matrix(rnorm(200 * 8), 200, 8) %*% chol(S * runif(1, 0.5, 2))
})
pair_d <- function(cl) {
  d <- c(); k <- length(cl)
  for (i in 1:(k - 1)) for (j in (i + 1):k) d <- c(d, riemann_distance(cl[[i]], cl[[j]]))
  d
}
raw <- pair_d(lapply(sessions, cov))
rc  <- pair_d(lapply(whiten_batch(sessions, mode = "recenter"), function(o) cov(o$Z)))
boxplot(list(raw = raw, recentered = rc), ylab = "pairwise Riemannian distance",
        col = c("#FDDBC7", "#2166AC"), main = "between-session distance")
```

**Euclidean alignment** (`euclidean_alignment()`) instead builds one shared
transform from a grand-mean covariance:

```{r ea}
ea <- euclidean_alignment(X_list, input = "raw", mean_method = "logeuclid")
str(ea$aligned, max.level = 1)
```

At predict time, `self_center = TRUE` centers test data on its own mean, which
helps when the test session has drifted relative to training:

```{r self-center}
m_tr <- whiten_model(X_list[[1]], method = "ZCA", lambda = 0.1)
Z_drift <- predict(m_tr, X_list[[2]], self_center = TRUE)
dim(Z_drift)
```

## Riemannian tangent space

The standard Riemannian BCI pipeline projects per-trial covariance matrices to
the tangent space at a reference mean and vectorizes them, so ordinary linear
classifiers can be used. `epoch_covariances()` builds the covariances from a
trial tensor; `tangent_space()` maps them.

```{r tangent}
X_tensor <- array(rnorm(40 * 8 * 128), dim = c(40, 8, 128))
covs <- epoch_covariances(X_tensor, lambda = 0.05)
V <- tangent_space(covs, mean_method = "riemann")
dim(V)                       # 40 trials x p(p+1)/2 features

covs_rec <- untangent_space(V)   # inverse map
max(abs(covs_rec[[1]] - covs[[1]]))
```

### Riemannian distances, means, and barycenter whitening

The geometry primitives behind alignment are available directly. `spd_mean()`
computes the geometric (Riemannian), log-Euclidean, or arithmetic mean of a set
of covariances; `riemann_distance()` gives the affine-invariant distance; and
`barycenter_whitener()` fits a whitener that maps the geometric mean of a set of
covariances to the identity.

```{r riemann}
M <- spd_mean(covs, metric = "riemann")
riemann_distance(covs[[1]], M)

bw <- barycenter_whitener(covs, metric = "riemann")
Z_bw <- predict(bw, matrix(rnorm(100 * 8), 100, 8))
dim(Z_bw)
```

`lambda = "auto"` also works with `shrink_target = "diagonal"` (a scale-invariant
Schaefer-Strimmer intensity), and with `"auto"` inside `epoch_covariances()` for
per-epoch shrinkage.

## Relative / generalized-eigenvalue whitening

`whiten_relative()` whitens one covariance by a reference covariance and
diagonalizes the target -- the building block of CSP and xDAWN. The returned
`values` are the per-component variance ratios.

```{r relative}
Sigma_signal <- cov(matrix(rnorm(200 * 6), 200, 6) %*% diag(c(4, 3, 2, 1, 1, 1)))
Sigma_noise  <- cov(matrix(rnorm(200 * 6), 200, 6))
gr <- whiten_relative(Sigma_signal, reference = Sigma_noise, n_comp = 3)
round(gr$values, 3)
```

## Online updates with forgetting

`whiten_model_update()` refreshes a model from summary moments. A `decay < 1`
down-weights past data to track non-stationary drift in streaming settings.

```{r update}
X_new <- matrix(rnorm(80 * 16), 80, 16)
m2 <- whiten_model_update(m, X_new, decay = 0.9)
m2$n_obs
```

## Tensor data

3D EEG tensors (trials x channels x time) are handled directly.

```{r tensor-whiten}
m_t <- whiten_model_tensor(X_tensor, method = "ZCA", lambda = 0)
Z_t <- predict_tensor(m_t, X_tensor)
dim(Z_t)
X_t_rec <- unwhiten_tensor(m_t, Z_t)
dim(X_t_rec)
```

## Reporting

```{r report}
txt <- report_whitening(m, data = X_train, file = NULL)
cat(substr(txt, 1, 280), "...\n")
```

## References

- He, H., and Wu, D. (2019). Transfer learning for brain-computer interfaces:
  A Euclidean space data alignment approach. *IEEE TBME*.
- Barachant, A., et al. (2012). Multiclass brain-computer interface
  classification by Riemannian geometry. *IEEE TBME*.
- Haufe, S., et al. (2014). On the interpretation of weight vectors of linear
  models in multivariate neuroimaging. *NeuroImage*.
